---
title: "Final Portfolio"
author: "Ella Addink"
format: html
embed-resources: true
code-tools: true 
---

```{r}
#| include: false

library(vegabrite)
library(dplyr)
library(tidyr)
library(jsonlite)
library(readxl)
library(lubridate)
```

## Exercise 1 - Twin Genetics

a.  An example with something I like

b.  An example with something I don't like

c.  My graphics + stories

```{r}
genetics_wide <- read.csv("https://calvin-data304.netlify.app/data/twins-genetics-wide.csv")
genetics_long <- read.csv("https://calvin-data304.netlify.app/data/twins-genetics-long.csv")
```

Comparing twins

```{r}
vl_chart() |> 
  vl_mark_bar() |> 
  vl_encode_x("kit:O") |>
  vl_scale_x(paddingInner = 0) |> 
  vl_encode_y("genetic.share:Q") |>
  vl_encode_color("region:N") |> 
  vl_encode_xOffset("twin:O") |> 
  vl_facet("pair:O", columns = 2, title = "") |>
  vl_add_data(genetics_long)
```

Comparing DNA kits

```{r}
vl_chart() |> 
  vl_mark_bar() |> 
  vl_encode_x("pair:O") |>
  vl_scale_x(paddingInner = 0) |> 
  vl_encode_y("genetic.share:Q") |>
  vl_encode_color("region:N") |> 
  vl_encode_xOffset("twin:O") |> 
  vl_facet("kit:N", columns = 1) |>
  vl_add_data(genetics_long)
```

## Exercise 2 - Data & Graphic Challenge

## Exercise 3 - Tanzania Graphic

[Tanzania data](Tanzania_data.xlsx)

```{r}
tanzania <- read_excel(path = 'Tanzania_data.xlsx') 
```

To do points and lines use mark line with point = TRUE

```{r}
contraception <- 
vl_chart() |> 
  vl_mark_line(point = TRUE) |>
  vl_encode_x("date:T", title = "") |> 
  vl_encode_y("contraception_use:Q") |> 
  vl_encode_color("fertility_rate:Q") |> 
  vl_add_data(tanzania) 

fam_planning <- 
  
```

## Exercise 4 - Masterpiece

## Exercise 5 - Using features

Where I used these features:

-   An encoding channel other than x or y:

-   Layers:

-   Facets:

-   Concatenation or repeat:

-   Non-default settings for a channel’s scale or guide:

-   Tooltips:

-   Another kind of interaction (panning/zooming, brushing, sliders, etc.):

## Exercise 6 - Learning

a.  Name 2 or 3 examples of where I used a feature we did not discuss in class (a new kind of mark, transform, a way to customize something, a way to use an interaction, etc)

b.  Name 2 or 3 examples of where I followed the advice of Wilke or Knaflic
