---
title: "Final Portfolio"
author: "Ella Addink"
format: html
embed-resources: true
code-tools: true 
---

```{r}
#| code-fold: true 
#| warning: false

library(vegabrite)
library(dplyr)
library(tidyr)
library(jsonlite)
library(readxl)
library(lubridate)
```

## Exercise 1 - Twin Genetics

a.  An example with something I like

b.  An example with something I don't like

c.  My graphics + stories

```{r}
#| code-fold: true 
genetics_wide <- read.csv("https://calvin-data304.netlify.app/data/twins-genetics-wide.csv")
genetics_long <- read.csv("https://calvin-data304.netlify.app/data/twins-genetics-long.csv")
```

Comparing twins

```{r}
#| code-fold: true 
vl_chart() |> 
  vl_mark_bar() |> 
  vl_encode_x("kit:O") |>
  vl_scale_x(paddingInner = 0) |> 
  vl_encode_y("genetic.share:Q") |>
  vl_encode_color("region:N") |> 
  vl_encode_xOffset("twin:O") |> 
  vl_facet("pair:O", columns = 2, title = "") |>
  vl_add_data(genetics_long)
```

Comparing DNA kits

```{r}
#| code-fold: true 
vl_chart() |> 
  vl_mark_bar() |> 
  vl_encode_x("pair:O") |>
  vl_scale_x(paddingInner = 0) |> 
  vl_encode_y("genetic.share:Q") |>
  vl_encode_color("region:N") |> 
  vl_encode_xOffset("twin:O") |> 
  vl_facet("kit:N", columns = 1) |>
  vl_add_data(genetics_long)
```

## Exercise 2 - Data & Graphic Challenge

## Exercise 3 - Tanzania Graphic

[Tanzania data](Tanzania_data.xlsx)

```{r}
#| code-fold: true 
tanzania <- read_excel('Tanzania_data.xlsx') |>
  mutate(contraception_use = contraception_use / 100, 
         fam_plan_unmet = fam_plan_unmet / 100)
```

```{r}
#| code-fold: true 
points_base <- vl_chart() |> 
  vl_encode_x("date:T", title = "") |>
  vl_mark_point(strokeWidth = 1) |> 
  vl_encode_fill(value = "lightgrey") |> 
  vl_encode_opacity(value = 1) |>
  vl_encode_color(value = "black") |> 
  vl_encode_size("fertility_rate:Q", 
                 title = "Total Fertility Rate",
                 legend = list(orient = "bottom")) |> 
  vl_scale_size(domainMin = 5, domainMax = 7, rangeMax = 170)

contraception_points <- points_base |>
  vl_encode_y("contraception_use:Q") |> 
  vl_axis_y(format = "%")

fam_plan_points <- points_base |> 
  vl_encode_y("fam_plan_unmet:Q")

contraception_line <- vl_chart() |> 
  vl_mark_line() |> 
  vl_encode_x("date:T", title = "") |> 
  vl_encode_y("contraception_use:Q", title = "") |>
  vl_encode_color(value = "#7570b3")
  
fam_plan_line <- vl_chart() |> 
  vl_mark_line() |> 
  vl_encode_x("date:T", title = "") |> 
  vl_encode_y("fam_plan_unmet:Q", title = "") |> 
  vl_encode_color(value = "#1b9e77") 

contraception_label <- vl_chart() |> 
  vl_mark_text(dx = 260, size = 12) |> 
  vl_encode_text(value = "Contraception Use") |> 
  vl_encode_y(datum = 0.384) |> 
  vl_encode_color(value = "#7570b3")

fam_plan_label <- vl_chart() |>
  vl_mark_text(dx = 288, size = 12) |>
  vl_encode_text(value = "Unmet Family Planning Need") |>
  vl_encode_y(datum = 0.221) |> 
  vl_encode_color(value = "#1b9e77")

vl_layer(fam_plan_line, contraception_line, 
         fam_plan_points, contraception_points, 
         contraception_label, fam_plan_label) |> 
  vl_add_data(tanzania) |>
  vl_add_properties(width = 400, height = 300, 
                    title = "Contraception Use and Family Planning Increase in Tanzania")
```

## Exercise 4 - Masterpiece

## Exercise 5 - Using features

Where I used these features:

-   An encoding channel other than x or y:

-   Layers:

-   Facets:

-   Concatenation or repeat:

-   Non-default settings for a channel’s scale or guide:

-   Tooltips:

-   Another kind of interaction (panning/zooming, brushing, sliders, etc.):

## Exercise 6 - Learning

a.  Name 2 or 3 examples of where I used a feature we did not discuss in class (a new kind of mark, transform, a way to customize something, a way to use an interaction, etc)

b.  Name 2 or 3 examples of where I followed the advice of Wilke or Knaflic
